Using raw accelerometer data to predict high-impact mechanical loading

### 2022-03-04 manuscript/manuscript.Rmd
Revise the manuscript

Bone reponds to mechanical loading by altering its properties, such as density, geometry, and strength [@Turner2003; @Fonseca2014; @Santos2017]. Physical activity is a way of eliciting mechanical stimuli, and it is well established to improve bone health. However, there is no consensus in which characteristics of the physical activity-related mechanical stimulus are better associated with bone health benefits. Nevertheless, there is evidence that that high-impact exercise, although performed scarcely throughout the day, may play an important role in inducing positive bone adaptations [@Vainionpaa2006; @Stiles2017].

But, even when knowing precisely which the best loading regimen is to improve bone health, the question of how to measure and monitor this load remains. Bone mechanical loading is usually measured in laboratory settings, using instruments such as force plates (FP) [@Medved2000]. Measurements in these instruments, that are confined to a laboratory space, are not applicable to controling the loads a person is routinely subjected in free-living contexts. To surpass this issue, in the last few years, methods using wearable devices such as accelerometers have been proposed to measure mechanical loading, more precisely,  ground reaction force (GRF) [@Neugebauer2012; @Neugebauer2014; @Fortune2014; @Neugebauer2018; @Veras2020] and loading rate (LR) [@Fortune2014; @Veras2020]. These devices are light, small, can be used throughout days, and have a relatively affordable price. The aforementioned characteristics make the accelerometers well suited instruments to measure mechanical loading on a daily basis.

However, the validity of the accelerometer-predicted mechanical loading values depends on some factors, as the sample characteristics [@Welk2005]. In fact, due to this, several calibration studies have been done in several populations, as children [@Neugebauer2012], adults [@Neugebauer2014; @Fortune2014], military personnel [@Neugebauer2018], and subjects with obesity [@Veras2020]. Another important factor that influences the predictions validity is the type of activities performed in the calibration study [@Bassett2012]. Until now, studies that calibrate accelerometers to predict mechanical loading variables have usually employed activities as walking [@Fortune2014; @Neugebauer2018; @Veras2020], or walking and running [@Neugebauer2012; @Neugebauer2014]. Nevertheless, activities eliciting higher impact loads have not yet been tested.

Therefore, the objective of this study was to develop accelerometry-based mechanical loading prediction equations for high-impact activites in subjects from normal weight to severe obesity.

@@

In this study, jumps were chosen as the high-impact activites to be performed. After familiarization, participants completed a series of trials envolving jumps of different types and heights. First, drop jumps were executed from steps varying from 5cm to 40cm high, with 5cm increments. Then, box jumps were performed from the floor to 5cm, 15cm and 30cm high boxes. Finally, continuous jumps were carried out in two different heights: 5cm and 15cm. Drop and box jumps were executed in two sets of four jumps at each height, with 30s interval, and the continuous jumps in one set of 20s at each height. The landing phase of the drop jumps, the take-off phase of the box jumps and both phases of the continuous jumps were recorded in FC (1000Hz; AMTI Corporation, Watertown, USA). Participants were instructed to jump with no restrictions to arms movement, land two-footed and then remain stationary on the force plate for five seconds (except in the continuous jumps).

@@

The MATLAB software (Version 2019a, Mathworks, Natick, USA) was used to process data from both FP and accelerometers. Data were processed according to the following procedures: first, the GRF signal was resampled to match the activity monitors sampling frequency (100Hz). Then, both GRF and ACC signals were filtered using a Butterworth fourth-order low-pass filter, with 20 Hz cut-off frequency, to attenuate the noise magnitude. This cut-off frequency was selected based on visual inspection of the signals frequency spectrum after a Fast Fourier Transform (FFT). Afterwards, the GRF and ACC resultant vector was calculated ($r_i = \sqrt{x_i^2 + y_i^2 + z_i^2}$). After that, both signals were adjusted using the time set by the systems clock and then synchronized by the maximum cross-correlation coefficient, with posterior manual adjustments if necessary. Next, peak ACC (pACC) were determined by the following criteria: i) a minimum height of three standard deviations above the mean ACC value recorded during a trial; and ii) a separation of at least 0.2 seconds for continuous jumps and 4 seconds for the other jumps. Peak GRF (pGRF) were then defined as the highest value between 0.2 or 4 seconds before and after each pACC, according to the jump type. The rates of change were computed through a centered derivative from the beginning of the foot contact to the curve peak (Equation 1). The peak LR (pLR) and peak acceleration rate (pAR) were defined as the maximum value in the array (Equation 2).

@@

Linear mixed models were used to develop pGRF and pLR prediction equations for each vector (resultant and vertical) and accelerometer placement (ankle, lower back and  hip). Body mass was tested as predictor in all models, along with the acceleration magnitude (pACC) in the pGRF models and its rate (pAR) in the pLR models. These predictors were entered as fixed effects and kept in the model, as all have shown to be significant (*p* < 0.05). Subject and the interaction between jump type and height were tested as random effects, and both factors have shown to improve the models. Final models were chosen according to the -2 log-likelihood statistic [@Field2012]. The conditional coefficient of determination (R^2^), which estimates the variance explained by both fixed and random factors [@Nakagawa2013], was computed.

@@

Lastly, our pGRF and pLR prediction models were compared with a previously published equation by our group [@Veras2020]. As, for the best of our knowledge, there is no mechanical loading prediction equation developed for jumping activies, we chose to compare it with previous models from our group that were validated in a similar sample. The models used in the comparison were the ones predicting pGRF and pLR from both the resultant and vertical vectors of accelerometers worn at lower back and hip. The MAE, MAPE and RMSE were calculated using the values predicted using Veras et al. [@Veras2020] equation.

@@

In our protocol, three distinct jump types were performed, in several different heights. This variety of jump trials produced a varying response in both the FP- and accelerometer-measured variables. These responses in pGRF, pLR, pACC and pAR, regarding their resultant vector, can be observed in Figure 1 (see Supplemental Figure S1 for the vertical vector). Panels A and B show the FP-measured variables normalized by the subjects body mass, the pGRF and pLR, respectively. It can be seen that in the drop jump trials, there is a tendency to increase the loading magnitude the higher the jump, and in the box jumps this tendency is lower. As for the continuous jumps, both assayed heights elicit resonably constant loadings. The same pattern can be observed in the accelerometer-measured variables, pACC and pAR, in Panels C and D, respectively. Also, regarding the influence of the accelerometer placement, in most of the trials accelerometers worn at hip and lower back produced approximately similar values, with a tendency to higher dispersion in data derived from lower back accelerometers.

@@

Accelerometer-based mechanical loading prediction equations were developed based on linear mixed models. The regression coefficients, R^2^ and accuracy indices of the models derived from accelerometers worn at ankle, lower back and hip, and from the resultant and vertical vectors are presented in Table 1. The R^2^ values for the pGRF models ranged from `r min(GRF_accuracy$R2_cond)` to `r max(GRF_accuracy$R2_cond)`, showing that the models can explain at least `r round(min(GRF_accuracy$R2_cond), 1) * 100`\% of the pGRF variance. The observed MAPE values were between `r myround(min(GRF_accuracy$MAPE) * 100, 1)`\% and `r myround(max(GRF_accuracy$MAPE) * 100, 1)`\%, which falls outside of the proposed threshold of 8\%. Nevertheless, these MAPE values are similar to the results usually found in accelerometer-based energy expenditure prediction [@Diniz-Sousa2020], which have widespread use in research. Despite similar accuracy results achieved by all equations, data from the resultant vector of accelerometers at the hip have shown the lower accuracy error, represented by the smallest MAE (`r myround(hip_GRF_accuracy$MAE, 1)` N), MAPE (`r myround(hip_GRF_accuracy$MAPE * 100, 1)`\%) and RMSE (`r myround(hip_GRF_accuracy$RMSE, 1)` N). Compared to the pGRF models, the pLR equations yielded poorer results, which is expected based on past research [@Fortune2014; @Veras2020]. Their R^2^ ranged from `r min(LR_accuracy$R2_cond)` to `r max(LR_accuracy$R2_cond)` and MAPE from `r myround(min(LR_accuracy$MAPE) * 100, 1)`\% to `r myround(max(LR_accuracy$MAPE) * 100, 1)`\%. However, similar to the pGRF models, the best pLR prediction equation in terms of prediction accuracy was the one from the resultant vector of hip-worn accelerometers, with a MAE of `r myround(hip_LR_accuracy$MAE, 1)` N, MAPE of `r myround(hip_LR_accuracy$MAPE * 100, 1)`\% and RMSE of `r myround(hip_LR_accuracy$RMSE, 1)` N.

@@

To visualize the agreement between the pGRF and pLR values measured by the FP and predicted by the models, a series of Bland-Altman plots were built for both the resultant and vetical vectors and for all accelerometer placements. Figure 2 shows the Bland-Altman plots for the pGRF (Panel A) and pLR (Panel B) resultant vector of hip-worn accelerometers. Both plots have shown good agreement levels, as at least `r ba_agreement`\% of the data points are within the limits of agreement and data tended to cluster around zero. Bland-Altman plots for the other accelerometer placements and vectors have shown the same patterns described previously, and can be found in the Supplemental Figure S2. Also, as could be observed by the one-sample *t*-tests results, none of the equations presented a bias different than zero (*p* > 0.05). Furthermore, all of the equations have shown a proportional bias, i.e., the magnitude of the values influence the prediction error. However, as the linear regressions presented a very small R^2^ (maximum value of `r myround(max(ba_regression_ankle_R2), 2)`) this effect is neglectable.

@@

Although the accuracy of our equations showed to be beyond the established 8\% threshold, when applying a model not developed for jumping activies we can se that the accuracy tremendously decreases, reaching values of MAPE around `r floor(mean(as.numeric(substr(tab2_df$MAPE, 1, nchar(tab2_df$MAPE) - 2))[1:4]) / 50) * 50`\% for pGRF prediction and around `r floor(mean(as.numeric(substr(tab2_df$MAPE, 1, nchar(tab2_df$MAPE) - 2))[5:8]) / 10) * 10`\% for pLR prediction as can be observed in Table 2. Furthermore, applying to our accelerometer-recorded jumping data a model that is not specific for this activity yield unresonable results, such as the negative values for pGRF that can be observed in Figure 3, Panel A.

Finally, it is important to mention that an R package was developed [@Veras2021] to make easier to apply all the necessary data processing steps to use the prediction equations. The package website (https://lveras.com/impactr) has all the installation instructions and documentation necessary to apply this method on your own data.

@@

This study aimed to develop accelerometer-based mechanical loading predition equations for jumps of diverse types and heights, testing different accelerometer placements and using both the resultant and vertical vectors. The results have shown that jumping pGRF prediction models based on accelerometry data have a lower accuracy compared to models developed for other activies, such as walking and running. Despite that, these models are the correct choice for this kind of activity eliciting hight levels of impact, as applying a model developed for lower impact activities in a sample similar to the current study yielded far worse results in terms of prediction accuracy. Also, as expected, the pLR prediction models performed poorly compared to the pGRF equations.

Bone is known to increase its tissue mass and enhance its micro-architecture arrangement in response to mechanical stimulation [@Turner2003; @Fonseca2014], particularly at higher magnitudes [@Vainionpaa2006; @Stiles2017]. This type of high-intensity impact activies have shown to be more associated with bone health improval, even if they are usually performed in low quantities throughout the day. In fact, Vainionpää et al [@Vainionpaa2006] demonstrated that performing a relatively low number (<100) of pACC above 4.9*g* daily is positively associated with increases in bone mineral density. Usually, this acceleration magnitude can be achieved mostly by ballistic movements such as jumps [@Vainionpaa2006; @Rowlands2012]. Nevertheless, accelerometry-based mechanical loading regression models have mostly been validated with a set of activies, such as walking and running, that elict relatively low acceleration magnitudes, usually ranging from approximately 1.2 to 6*g* [@Neugebauer2014; @Veras2020].

@@

Loading rate is a highly sensitive parameter, that is usually calculated by deriving the force from the beggining of the foot contact to the floor to the point where the pGRF is achieved [@Liikavainio2007; @Fortune2014; @Veras2020]. This is a small time window, usually less than 0.2 seconds for the jumps that were performed in our protocol, and requires a fine sampling acquisition that captures small graduations on the force values throughout time, which is difficult to achieve with common accelerometer sampling frequecies, that are usually 100Hz at maximum. Based on this and on on results from previous investigations [@Fortune2014; @Veras2020], pLR prediction models are expected to have a lower accuracy compared to the pGRF equations. Nevertheless, although our pLR prediction models had a worse accuracy compared with the pGRF models, its accuracy was similar or even better than the results from previous studies that developed models to predict pLR during walking only [@Fortune2014; @Veras2020].

In our protocol, three accelerometer placements (ankle, lower back and hip) and two vectors (resultant and vertical) have been tested. Overall, the prediction models performance was reasonably consistent among the different placement and vector combinations, however, some conclusions can be draw. First, the hip accelerometer placement, that has traditionally been widely employed in research for the direct measure of physical activity [@Migueles2017], yielded a sligthly better accuracy compared the other placements tested. This could be useful, as previous data collected from hip-worn accelerometers, even for other purposes, can be adequately used to obtain mechanical loading predictions. Also, studies that aim to extract mechanical loading and other variables from accelerometry data, that use the hip as the default placement, can comfortably do so. Second, as the magnitude of the resultant vector of both pGRF and pLR, in the activities that were tested, is derived mostly from the vertical vector, and there were no considerable differences in performance between models using both vectors, they can be used interchangeably to predict mechanical loading. Besides that, while models using data from the vertical vector allow the use uniaxial accelerometers, data derived from the resultant vector have the important advantage of not depending on the correct acceleromter axes orientation, that cannot be guaranteed in free-living data collection [@Welk2005].

The development of accelerometry-based mechanical loading prediction models was done with the direct purpose of allowing to assess the loading beared by the skeletal system during daily physical activity. This would allow some clinical applications. For instance, these models, when used in combination with other prediction equations developed for lower impact activities [@Veras2020], can be used to fully analyse the mechanical loading spectrum that people with different characteristics are submitted to daily and relate to bone health parameters. This would allow a more complete understanding on how mechanical loading influence bone healh, and also allow to identify which loading volume and magnitude range, for example, are more efficient in inducing bone health benefits. Also, by the development and distribution of an R package that automates the steps necessary to apply the models that were proposed was made to enhance its applicability [@Veras2021]. This can prove to be an important tool, as the application of these models depend on non-trivial data processing skills, such as applying digital filters to reduce noise from the signal, detection of the peaks in this signal and the computation of derivatives to obtain the acceleration rate, necessary for the prediction of pLR.

@@

Mechanical loading prediction equations based on accelerometer data have already been proposed for several populations [@Neugebauer2012; @Neugebauer2018; @Veras2020] and using accelerometer in different body placements [@Fortune2014; @Veras2020]. What these past few studies have in common is that they validated their models using locomotion activities, as walking only or walking and runnning. The current study, by setting means to predict mechanical loading in several types and heights of jumps, broadens the application of such models, getting a step closer of the ability to accurately identify daily-living mechanical loading due to physical activity. This ability would allow the monitoring of loading in different contexts and to more fully understand the relationship between mechanical stimulation and its consequent effect on bone health.

>>>>

Bone responds to mechanical loading by altering its properties, such as density, geometry, and strength [@Turner2003; @Fonseca2014; @Santos2017]. Physical activity is a way of eliciting mechanical stimuli, and it is well established to improve bone health. However, there is no consensus in which characteristics of the physical activity-related mechanical stimulus are better associated with bone health benefits. Nevertheless, there is evidence that high-impact exercise, although performed scarcely throughout the day, may play an important role in inducing positive bone adaptations [@Vainionpaa2006; @Stiles2017].

But, even when knowing precisely which the best loading regimen is to improve bone health, the question of how to measure and monitor this load remains. Bone mechanical loading is usually measured in laboratory settings, using instruments such as force plates (FP) [@Medved2000]. Measurements in these instruments, that are confined to a laboratory space, are not applicable in controlling the loads that a person is routinely subjected to in free-living contexts. To surpass this issue, in the last few years, methods using wearable devices such as accelerometers have been proposed to measure mechanical loading, more precisely,  ground reaction force (GRF) [@Neugebauer2012; @Neugebauer2014; @Fortune2014; @Neugebauer2018; @Veras2020] and loading rate (LR) [@Fortune2014; @Veras2020]. These devices are light, small, can be used throughout days, and have a relatively affordable price. These characteristics make the accelerometers well-suited instruments to measure mechanical loading on a daily basis.

However, the validity of the accelerometer-predicted mechanical loading values depends on some factors, as the sample characteristics [@Welk2005]. In fact, due to this, several calibration studies have been done in several populations, such as children [@Neugebauer2012], adults [@Neugebauer2014; @Fortune2014], military personnel [@Neugebauer2018], and subjects with obesity [@Veras2020]. Another important factor that influences the validity of the predictions is the type of activities performed in the calibration study [@Bassett2012]. Until now, studies that calibrate accelerometers to predict mechanical loading variables have usually employed activities such as walking [@Fortune2014; @Neugebauer2018; @Veras2020], or walking and running [@Neugebauer2012; @Neugebauer2014]. Nevertheless, activities eliciting higher impact loads have not yet been tested.

Therefore, the objective of this study was to develop accelerometry-based mechanical loading prediction equations for high-impact activities in subjects from normal weight to severe obesity.

@@

In this study, jumps were chosen as the high-impact activities to be performed. After familiarization, participants completed a series of trials involving jumps of different types and heights. First, drop jumps were executed from steps varying from 5cm to 40cm high, with 5cm increments. Then, box jumps were performed from the floor to 5cm, 15cm, and 30cm high boxes. Finally, continuous jumps were carried out at two different heights: 5cm and 15cm. Drop and box jumps were executed in two sets of four jumps at each height, with 30s interval, and the continuous jumps in one set of 20s at each height. The landing phase of the drop jumps, the take-off phase of the box jumps, and both phases of the continuous jumps were recorded in FC (1000Hz; AMTI Corporation, Watertown, USA). Participants were instructed to jump with no restrictions to arms movement, land two-footed, and then remain stationary on the force plate for five seconds (except in the continuous jumps).

@@

The MATLAB software (Version 2019a, Mathworks, Natick, USA) was used to process data from both FP and accelerometers. Data were processed according to the following procedures: first, the GRF signal was resampled to match the activity monitors sampling frequency (100Hz). Then, both GRF and ACC signals were filtered using a Butterworth fourth-order low-pass filter, with 20 Hz cut-off frequency, to attenuate the noise magnitude. This cut-off frequency was selected based on visual inspection of the signals frequency spectrum after a Fast Fourier Transform (FFT). Afterwards, the GRF and ACC resultant vector was calculated ($r_i = \sqrt{x_i^2 + y_i^2 + z_i^2}$). After that, both signals were adjusted using the time set by the systems clock and then synchronized by the maximum cross-correlation coefficient, with posterior manual adjustments if necessary. Next, peak ACC (pACC) was determined by the following criteria: i) a minimum height of three standard deviations above the mean ACC value recorded during a trial; and ii) a separation of at least 0.2 seconds for continuous jumps and 4 seconds for the other jumps. Peak GRF (pGRF) was then defined as the highest value between 0.2 or 4 seconds before and after each pACC, according to the jump type. The rates of change were computed through a centered derivative from the beginning of the foot contact to the curve peak (Equation 1). The peak LR (pLR) and peak acceleration rate (pAR) were defined as the maximum value in the array (Equation 2).

@@

Linear mixed models were used to develop pGRF and pLR prediction equations for each vector (resultant and vertical) and accelerometer placement (ankle, lower back, and  hip). Body mass was tested as a predictor in all models, along with the acceleration magnitude (pACC) in the pGRF models and its rate (pAR) in the pLR models. These predictors were entered as fixed effects and kept in the model, as all have shown to be significant (*p* < 0.05). Subject and the interaction between jump type and height were tested as random effects, and both factors have shown to improve the models. Final models were chosen according to the -2 log-likelihood statistic [@Field2012]. The conditional coefficient of determination (R^2^), which estimates the variance explained by both fixed and random factors [@Nakagawa2013], was computed.

@@

Lastly, our pGRF and pLR prediction models were compared with a previously published equation by our group [@Veras2020]. As, for the best of our knowledge, there is no mechanical loading prediction equation developed for jumping activities, we chose to compare it with previous models from our group that were validated in a similar sample. The models used in the comparison were the ones predicting pGRF and pLR from both the resultant and vertical vectors of accelerometers worn at lower back and hip. The MAE, MAPE and RMSE were calculated using the values predicted using Veras et al. [@Veras2020] equation.

@@

In our protocol, three distinct jump types were performed, in several different heights. This variety of jump trials produced a varying response in both the FP- and accelerometer-measured variables. These responses in pGRF, pLR, pACC, and pAR, regarding their resultant vector, can be observed in Figure 1 (see Supplemental Figure S1 for the vertical vector). Panels A and B show the FP-measured variables normalized by the subjects body mass, the pGRF and pLR, respectively. It can be seen that in the drop jump trials, there is a tendency to increase the loading magnitude the higher the jump, and in the box jumps this tendency is lower. As for the continuous jumps, both assayed heights elicit reasonably constant loadings. The same pattern can be observed in the accelerometer-measured variables, pACC and pAR, in Panels C and D, respectively. Also, regarding the influence of the accelerometer placement, in most of the trials accelerometers worn at hip and lower back produced approximately similar values, with a tendency to higher dispersion in data derived from lower back accelerometers.

@@

Accelerometer-based mechanical loading prediction equations were developed based on linear mixed models. The regression coefficients, R^2^ and accuracy indices of the models derived from accelerometers worn at ankle, lower back, and hip, and from the resultant and vertical vectors are presented in Table 1. The R^2^ values for the pGRF models ranged from `r min(GRF_accuracy$R2_cond)` to `r max(GRF_accuracy$R2_cond)`, showing that the models can explain at least `r round(min(GRF_accuracy$R2_cond), 1) * 100`\% of the pGRF variance. The observed MAPE values were between `r myround(min(GRF_accuracy$MAPE) * 100, 1)`\% and `r myround(max(GRF_accuracy$MAPE) * 100, 1)`\%, which falls outside of the proposed threshold of 8\%. Nevertheless, these MAPE values are similar to the results usually found in accelerometer-based energy expenditure prediction [@Diniz-Sousa2020], which have widespread use in research. Despite similar accuracy results achieved by all equations, data from the resultant vector of accelerometers at the hip have shown the lower accuracy error, represented by the smallest MAE (`r myround(hip_GRF_accuracy$MAE, 1)` N), MAPE (`r myround(hip_GRF_accuracy$MAPE * 100, 1)`\%), and RMSE (`r myround(hip_GRF_accuracy$RMSE, 1)` N). Compared to the pGRF models, the pLR equations yielded poorer results, which is expected based on past research [@Fortune2014; @Veras2020]. Their R^2^ ranged from `r min(LR_accuracy$R2_cond)` to `r max(LR_accuracy$R2_cond)` and MAPE from `r myround(min(LR_accuracy$MAPE) * 100, 1)`\% to `r myround(max(LR_accuracy$MAPE) * 100, 1)`\%. However, similar to the pGRF models, the best pLR prediction equation in terms of prediction accuracy was the one from the resultant vector of hip-worn accelerometers, with a MAE of `r myround(hip_LR_accuracy$MAE, 1)` N, MAPE of `r myround(hip_LR_accuracy$MAPE * 100, 1)`\%, and RMSE of `r myround(hip_LR_accuracy$RMSE, 1)` N.

@@

To visualize the agreement between the pGRF and pLR values measured by the FP and predicted by the models, a series of Bland-Altman plots were built for both the resultant and vertical vectors and for all accelerometer placements. Figure 2 shows the Bland-Altman plots for the pGRF (Panel A) and pLR (Panel B) resultant vector of hip-worn accelerometers. Both plots have shown good agreement levels, as at least `r ba_agreement`\% of the data points are within the limits of agreement and data tended to cluster around zero. Bland-Altman plots for the other accelerometer placements and vectors have shown the same patterns described previously, and can be found in the Supplemental Figure S2. Also, as could be observed by the one-sample *t*-tests results, none of the equations presented a bias different than zero (*p* > 0.05). Furthermore, all of the equations have shown a proportional bias, i.e., the magnitude of the values influences the prediction error. However, as the linear regressions presented a very small R^2^ (maximum value of `r myround(max(ba_regression_ankle_R2), 2)`) this effect is neglectable.

@@

Although the accuracy of our equations showed to be beyond the established 8\% threshold, when applying a model not developed for jumping activities, we can see that the accuracy tremendously decreases, reaching values of MAPE around `r floor(mean(as.numeric(substr(tab2_df$MAPE, 1, nchar(tab2_df$MAPE) - 2))[1:4]) / 50) * 50`\% for pGRF prediction and around `r floor(mean(as.numeric(substr(tab2_df$MAPE, 1, nchar(tab2_df$MAPE) - 2))[5:8]) / 10) * 10`\% for pLR prediction as can be observed in Table 2. Furthermore, applying to our accelerometer-recorded jumping data a model that is not specific for this activity yields unreasonable results, such as the negative values for pGRF that can be observed in Figure 3, Panel A.

Finally, it is important to mention that an R package was developed [@Veras2021] to make it easier to apply all the necessary data processing steps to use the prediction equations. The package website (https://lveras.com/impactr) has all the installation instructions and documentation necessary to apply this method on your own data.

@@

This study aimed to develop accelerometer-based mechanical loading prediction equations for jumps of diverse types and heights, testing different accelerometer placements and using both the resultant and vertical vectors. The results have shown that jumping pGRF prediction models based on accelerometry data have a lower accuracy compared to models developed for other activities, such as walking and running. Despite that, these models are the correct choice for this kind of activity eliciting hight levels of impact, as applying a model developed for lower impact activities in a sample similar to the current study yielded far worse results in terms of prediction accuracy. Also, as expected, the pLR prediction models performed poorly compared to the pGRF equations.

Bone is known to increase its tissue mass and enhance its micro-architecture arrangement in response to mechanical stimulation [@Turner2003; @Fonseca2014], particularly at higher magnitudes [@Vainionpaa2006; @Stiles2017]. This type of high-intensity impact activities have shown to be more associated with bone health improvements, even if they are usually performed in low quantities throughout the day. In fact, Vainionpää et al [@Vainionpaa2006] demonstrated that performing a relatively low number (<100) of pACC above 4.9*g* daily is positively associated with increases in bone mineral density. Usually, this acceleration magnitude can be achieved mostly by ballistic movements such as jumps [@Vainionpaa2006; @Rowlands2012]. Nevertheless, accelerometry-based mechanical loading regression models have mostly been validated with a set of activities, such as walking and running, that elicit relatively low acceleration magnitudes, usually ranging from approximately 1.2 to 6*g* [@Neugebauer2014; @Veras2020].

@@

Loading rate is a highly sensitive parameter, that is usually calculated by deriving the force from the beginning of the foot contact to the floor to the point where the pGRF is achieved [@Liikavainio2007; @Fortune2014; @Veras2020]. This is a small time window, usually less than 0.2 seconds for the jumps that were performed in our protocol, and requires a fine sampling acquisition that captures small graduations on the force values throughout time, which is difficult to achieve with common accelerometer sampling frequencies, that are usually 100Hz at maximum. Based on this and on results from previous investigations [@Fortune2014; @Veras2020], pLR prediction models are expected to have a lower accuracy compared to the pGRF equations. Nevertheless, although our pLR prediction models had a worse accuracy compared with the pGRF models, its accuracy was similar or even better than the results from previous studies that developed models to predict pLR during walking only [@Fortune2014; @Veras2020].

In our protocol, three accelerometer placements (ankle, lower back, and hip) and two vectors (resultant and vertical) have been tested. Overall, the prediction models performance was reasonably consistent among the different placement and vector combinations, however, some conclusions can be drawn. First, the hip accelerometer placement, that has traditionally been widely employed in research for the direct measure of physical activity [@Migueles2017], yielded a slightly better accuracy compared the other placements tested. This could be useful, as previous data collected from hip-worn accelerometers, even for other purposes, can be adequately used to obtain mechanical loading predictions. Also, studies that aim to extract mechanical loading and other variables from accelerometry data, that use the hip as the default placement, can comfortably do so. Second, as the magnitude of the resultant vector of both pGRF and pLR, in the activities that were tested, is derived mostly from the vertical vector, and there were no considerable differences in performance between models using both vectors, they can be used interchangeably to predict mechanical loading. Besides that, while models using data from the vertical vector allow the use of uniaxial accelerometers, data derived from the resultant vector have the important advantage of not depending on the correct accelerometer axes orientation, which cannot be guaranteed in free-living data collection [@Welk2005].

The development of accelerometry-based mechanical loading prediction models was done with the direct purpose of allowing to assess the loading beared by the skeletal system during daily physical activity. This would allow some clinical applications. For instance, these models, when used in combination with other prediction equations developed for lower impact activities [@Veras2020], can be used to fully analyse the mechanical loading spectrum that people with different characteristics are submitted to daily and relate to bone health parameters. This would allow a more complete understanding on how mechanical loading influences bone health, and also allow to identify which loading volume and magnitude range, for example, are more efficient in inducing bone health benefits. Also, by the development and distribution of an R package that automates the steps necessary to apply the models that were proposed was made to enhance its applicability [@Veras2021]. This can prove to be an important tool, as the application of these models depend on non-trivial data processing skills, such as applying digital filters to reduce noise from the signal, detection of the peaks in this signal and the computation of derivatives to obtain the acceleration rate, necessary for the prediction of pLR.

@@

Mechanical loading prediction equations based on accelerometer data have already been proposed for several populations [@Neugebauer2012; @Neugebauer2018; @Veras2020] and using an accelerometer in different body placements [@Fortune2014; @Veras2020]. What these past few studies have in common is that they validated their models using locomotion activities, such as walking only or walking and running. The current study, by setting means to predict mechanical loading in several types and heights of jumps, broadens the application of such models, getting a step closer to the ability to accurately identify daily-living mechanical loading due to physical activity. This ability would allow the monitoring of loading in different contexts and to more fully understand the relationship between mechanical stimulation and its consequent effect on bone health.
